---
title: "soil"
format: html
editor: visual
---

# Prep

## Setup

```{r setup, include=F, message=F,warning=F}
#| echo: false
#| cache: true
rm(list =ls())
#getwd()
#setwd("C:/Users/jadi0015/OneDrive - Umeå universitet/Documents/1_NissonCarbon/NissonCarbon")
library(readr)
library(readxl)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(brms)
library(tidybayes)
library(gt)
library(purrr)
library(gtExtras)
library(bayestestR)
library(emmeans)

```

## Data

# Plots

```{r}
soil_data <- readRDS("../1_data/final/soil_data.rds")

carbon_comp <- soil_data$carbon_comp
carbon_comp_list <- carbon_comp %>%
  arrange(veg, plot, treat, sample) %>% 
  group_split(veg,sample)


stocks <- soil_data$stocks
stock_list <- stocks %>%
  group_split(veg, sample)
```

## Depth

```{r sum_profile, warning=F, message=F}
#| echo: false
# ggplot(long_profile, aes(x = habitat,
#                    y = depth,
#                    fill = treat)) +
#   geom_boxplot() +
#   labs(x = "Habitat × sample",
#        y = "Layer Depth",
#        fill = "treat") +
#   scale_y_continuous() +
#   theme_bw() +
#   facet_wrap(~Layer)+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

## Bulk

```{r long profile, warning=F, message=F}
#| echo: false
# ggplot(carbon_comp, aes(x = veg,
#                    y = bulk,
#                    fill = treat)) +
#   geom_boxplot() +
#   labs(x = "Habitat × sample",
#        y = "Bulk_density",
#        fill = "treat") +
#   scale_y_continuous() +
#   theme_bw() +
#   facet_wrap(~sample)+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## carbon stock

test model str

```{r}
# model <- brm(
#     c_stock ~ treat + (1 | plot),
#     data = carbon_comp,
#     family = gaussian(),
#     chains = 4, iter = 8000, warmup = 2000, cores = 4,
#     silent = TRUE, 
#     refresh = 0,
#     control = list(adapt_delta = 0.95, max_treedepth = 15)
#   )
```

### total-brms

test \<- fit_one_model(l_stocks)

```{r model_brms, warning=FALSE,message=F}
#| echo: false
# options(brms.backend = NULL) <- NEVER PLAY WITH THIS AGAIN!
fit_model <- function(d, stock_var = "c_stock", quiet = TRUE) {

  veg_name  <- unique(d$veg)
  layer     <- unique(d$sample)

  # -----------------------
  # RAW SUMMARY
  # -----------------------
  raw_summary <- d %>%
    
    # mean and sd for each treatment
    group_by(treat) %>%
    summarise(
      raw_mean = mean(.data[[stock_var]], na.rm = TRUE),
      raw_sd   = sd(.data[[stock_var]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    
    # Add total (across all treatments) 
    # total being litter + humues
    bind_rows(
      d %>%
        summarise(
          treat    = "total",
          raw_mean = mean(.data[[stock_var]], na.rm = TRUE),
          raw_sd   = sd(.data[[stock_var]], na.rm = TRUE)
        )
    ) %>%
    pivot_wider(
      names_from = treat,
      values_from = c(raw_mean, raw_sd)
    ) %>%
    rename(
      ctl_mean   = raw_mean_C,
      ctl_sd     = raw_sd_C,
      sf_mean    = raw_mean_S,
      sf_sd      = raw_sd_S,
      total_mean = raw_mean_total,
      total_sd   = raw_sd_total
    )


  # -----------------------
  # MODEL FORMULA
  # -----------------------
  
  # used to have varying stock_vars
  formula <- as.formula(paste0(stock_var, " ~ treat + (1 | plot)"))

  # -----------------------
  # FIT MODEL
  # -----------------------
  model <- brm(
    formula = formula,
    data = d,
    family = gaussian(), # assume normal
    chains = 2, cores = 2, # rerun later with 4x4
    iter = 8000, warmup = 2000, # increase to improve "convergence"
    backend = "rstan",
    control = list(adapt_delta = 0.95, max_treedepth = 15),
    refresh = if (quiet) 0 else 100
  )

  # -----------------------
  # DIAGNOSTICS
  # -----------------------
  
  # summarise diagnoistics in one table
  diag <- tibble(
    veg = veg_name,
    layer = layer,
    stock = stock_var,
    n_divergent = sum(model$fit@sim$sampler_diagnostics[, "divergent__"]),
    max_Rhat     = max(summary(model)$fixed[, "Rhat"]),
    min_bulk_ESS = min(summary(model)$fixed[, "Bulk_ESS"]),
    min_tail_ESS = min(summary(model)$fixed[, "Tail_ESS"])
  )

  # -----------------------
  # POSTERIOR DRAWS
  # -----------------------
  
  # collect posteriors in one table
  draws <- spread_draws(model, b_Intercept, b_treatS) %>%
    mutate(
      control = b_Intercept,
      treat   = b_Intercept + b_treatS
    )

  posterior_long <- draws %>%
    select(.draw, control, treat) %>%
    pivot_longer(c(control, treat),
                 names_to = "treat",
                 values_to = "value") %>%
    mutate(
      veg = veg_name,
      layer = layer,
      stock = stock_var
    )

  # -----------------------
  # MARGINAL POSTERIOR MEANS (E[y])
  # -----------------------
  
  # summarise predictions for plots
  epred_summary <- posterior_long %>%
    group_by(treat, veg, layer, stock) %>%
    median_qi(value, .width = 0.95)


  # -----------------------
  # POSTERIOR SUMMARY TABLE
  # -----------------------
  
  # collect summary of data
  post_summary <- draws %>%
    summarise(
      post_mean_diff = mean(treat - control),
      lower_cred     = quantile(treat - control, 0.025),
      upper_cred     = quantile(treat - control, 0.975),
      prob_increase  = mean(treat > control),
      prob_decrease  = mean(treat < control),
      bf             = prob_increase / prob_decrease
    )
  
  # posterior difference vector
  diff_vals <- draws$treat - draws$control

  # Compute pd
  pd_val   <- pd(diff_vals)$pd
  
  
  # compute ROPE
  sd_diff <- sd(diff_vals)
  rope_bounds <- c(-1, 1) * sd_diff
  rope_val <- rope(diff_vals, range = rope_bounds)


  # bind results
  results <- bind_cols(raw_summary, post_summary) %>%
    mutate(
      veg = veg_name,
      layer = layer,
      stock = stock_var,
      mean = sprintf("%.4f ± %.4f", total_mean, total_sd),
      control   = sprintf("%.4f ± %.4f", ctl_mean, ctl_sd),
      snowfence = sprintf("%.4f ± %.4f", sf_mean, sf_sd),
      mean_diff = sprintf("%.4f", post_mean_diff),
      CI_low        = sprintf("%.4f", lower_cred),
      CI_high       = sprintf("%.4f", upper_cred),
      pd = sprintf("%.3f",pd_val),
      rope_pct = sprintf("%.4f", rope_val)
    ) %>%
    select(veg, layer, stock, control, snowfence,
           mean_diff, CI_low, CI_high,, pd, rope_pct)

  # -----------------------
  # RETURN RESULTS
  # -----------------------
  list(
    posterior_long = posterior_long,
    epred_summary  = epred_summary,
    diagnostics    = diag,
    results        = results
  )
}

# -----------------------
# SHOW RESULTS FUNCTIONS
# -----------------------

# view_results <- function(all_results){
#   results <- bind_rows(
#   lapply(seq_along(all_results), function(i) {
#     all_results[[i]]$results %>% mutate(dataset = i)
#     })
#   )
#   View(results)
# }
# view_diagnostics <- function(all_results){
#   results <- bind_rows(
#   lapply(seq_along(all_results), function(i) {
#     all_results[[i]]$diagnostics %>% mutate(dataset = i)
#     })
#   )
#   View(results)
# }
```

Run models

```{r, message = F, warning=F}
bulk_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "bulk")
})

depth_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "depth")
})

o_c_percent_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "o_c_percent")
})
o_n_percent_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "o_n_percent")
})
c_stock_results <- lapply(stock_list, function(d) {
  fit_model(d, stock_var = "c_stock")
})
n_stock_results <- lapply(stock_list, function(d) {
  fit_model(d, stock_var = "n_stock")
})
alkyl_stock_results <- lapply(stock_list, function(d) {
  fit_model(d, stock_var = "alkyl_stock")
})
o_alkyl_stock_results <- lapply(stock_list, function(d) {
  fit_model(d, stock_var = "o_alkyl_stock")
})
c_n_ratio_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "c_n_ratio")
})
d13c_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "d13c")
})
d15n_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "d15n")
})
alkyl_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "Alkyl")
})
carbohydrate_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "Carbohydrate")
})
carbonyl_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "carbonyl")
})
o_aromatic_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "o_aromatic")
})
aromatic_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "aromatic")
})
methoxy_n_alkyl_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "methoxy_n_alkyl")
})
o_alkyl_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "o_alkyl")
})
di_o_alkyl_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "di_o_alkyl")
})
Alkyl_o_alkyl_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "Alkyl/Carbohydrate")
})
Carbohydrate_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "Carbohydrate")
})
F_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "F")
})
G_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "G")
})
dS_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "dS")
})
eS_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "eS")
})
total_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "total_vascular")
})
alkyl_c_results <- lapply(carbon_comp_list, function(d) {
  fit_model(d, stock_var = "alkyl_c")
})

```

view results

```{r}
view_results(alkyl_c_results)
view_diagnostics(alkyl_c_results)
```

collect results

```{r}
# Get all object names ending with "_results"
result_names <- ls(pattern = "_results$")

# Put them into a named list
result_list <- mget(result_names)
```

Rerun single variables and add them

```{r}
# result_list <- readRDS("C:/Users/jadi0015/OneDrive - Umeå universitet/Documents/1_NissonCarbon/NissonCarbon/1_data/Model/all_results.rds")

# result_list[["Alkyl_o_alkyl_results"]] <- Alkyl_o_alkyl_results

# result_list[["total_results"]] <- total_results

# result_list[["alkyl_c_results"]] <- alkyl_c_results

```

save results

```{r}
# Save to RDS
saveRDS(result_list, 
        file = "C:/Users/jadi0015/OneDrive - Umeå universitet/Documents/1_NissonCarbon/NissonCarbon/1_data/Model/all_results.rds")
```
