---
title: "species_adaptation"
author: "Liyenne"
date: "2025-10-22"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("outputs")
    )
  })
cache: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup
```{r}
library(openxlsx)

# data handling
library(tidyverse)
library(broom)
library(purrr)

# visualization
library(ggplot2)

library(traitstrap)

# modelling
library(vegan)
library(lme4)
library(lmerTest)
library(glmmTMB)
library(pscl)
library(brms)

# model coefficients
library(emmeans)
library(MuMIn)

# model diagnostics
library(DHARMa)
library(easystats)
library(fitdistrplus)


library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

# Data import
```{r}
# taxonomic information
spec_list <- read_csv("2_data/1_raw/community/2025/sf_vp_species_list.csv") %>% 
  mutate(across(1:5, as.factor))

plot_data <- read.xlsx("2_data/1_raw/community/2025/SF_plot_data_2025.xlsx") %>% 
  mutate(across(plot:treatment, as.factor)) %>% 
  select(plot:treatment)
```

```{r data import}
traits_no_chem <- read_csv("2_data/2_clean/traits/traits_no_chem_w.csv") %>% 
  mutate(across(1:10, as.factor))

traits <- read_csv("2_data/2_clean/traits/traits_w.csv") %>% 
  mutate(across(1:10, as.factor))

traits_l <- read_csv("2_data/2_clean/traits/traits_l.csv") %>% 
  mutate(across(1:11, as.factor))

boot_traits <- read_csv("2_data/2_clean/traits/bootstrapped_traits_bmWeighted.csv") %>% 
  mutate(across(1:6, as.factor))

# global database
try_traits <- read_csv("2_data/2_clean/traits/try_traits.csv") %>% 
  select(-...1) 
  
# community composition data (l)
community <- read_csv("2_data/2_clean/sf_community.csv") %>% 
  mutate(across(1:10, as.factor))
```


# Analysis
## turnover vs. ITV (Leps et al., 2011)
### CWM calculations
#### fixed = landscape
```{r}
spec_traits_bs <- read.csv("2_data/2_clean/traits/filled_traits_bmWeighted.csv") %>% 
  group_by(spec_comm, trait) %>% 
  summarise(trait_value = mean(trait_value)) %>% 
  rename(spec = spec_comm) %>% 
  pivot_wider(names_from = trait, values_from = trait_value)
  
fix_avg <- community %>% 
  filter(part == "leaf") %>% 
  left_join(spec_traits_bs) %>% 
  mutate(height = biomass * height, 
            LDMC = biomass * LDMC, 
            SLA = biomass * SLA, 
            wN = biomass * wN, 
            d15N = biomass * d15N, 
            wC = biomass * wC, 
            d13C = biomass * d13C, 
            CN_ratio = biomass * CN_ratio) %>% 
  group_by(plot) %>% 
  summarise(biomass = sum(biomass, na.rm = TRUE), 
            height = sum(height, na.rm = TRUE), 
            LDMC = sum(LDMC, na.rm = TRUE), 
            SLA = sum(SLA, na.rm = TRUE), 
            wN = sum(wN, na.rm = TRUE), 
            d15N = sum(d15N, na.rm = TRUE), 
            wC = sum(wC, na.rm = TRUE), 
            d13C = sum(d13C, na.rm = TRUE), 
            CN_ratio = sum(CN_ratio, na.rm = TRUE)) %>% 
  mutate(height = height / biomass, 
            LDMC = LDMC / biomass, 
            SLA = SLA / biomass, 
            wN = wN / biomass, 
            d15N = d15N / biomass, 
            wC = wC / biomass, 
            d13C = d13C / biomass, 
            CN_ratio = CN_ratio / biomass, 
         var_type = "landscape")

# use bootstrapped CWMs for specific averages
spec_avg <- read.csv("2_data/2_clean/traits/bootstrapped_traits_bmWeighted.csv") %>% 
  group_by(plot, trait) %>% 
  summarise(mean = mean(mean)) %>% 
  pivot_wider(names_from = trait, values_from = mean) %>% 
  select(!c(dweight, fweight, leafa)) %>% 
  mutate(var_type = "local")
```

```{r}
ITV <- bind_rows(fix_avg, spec_avg) %>% 
  select(!biomass) %>% 
  pivot_longer(cols = height:CN_ratio, names_to = "trait", values_to = "trait_value") %>% 
  pivot_wider(names_from = var_type, 
              values_from = trait_value) %>% 
  mutate(ITV = local - landscape) %>% 
  pivot_longer(cols = landscape:ITV, names_to = "var_type", values_to = "CWM") %>% 
  mutate(habitat = as.factor(ifelse( # create habitat variable based on plot code
    grepl('D', plot), 'fen', ifelse(
      grepl('E', plot), 'heath', 'ridge'))
    )) %>% 
  mutate(treatment = as.factor(ifelse( # create treatment variable based on plot code
    grepl('C', plot), 'control', 'snowfence'
    ))) %>%  
  mutate(site_id = as.factor(substr(plot, 1,2))) %>%  # create site variable
  select(plot, site_id, habitat, treatment, everything())
```

```{r}
ITV_w <- ITV %>% 
  pivot_wider(names_from = var_type, values_from = CWM)

ITV %>% 
  filter(var_type == "local") %>% 
  ggplot(aes(x = habitat, y = CWM, fill = treatment)) + 
  geom_boxplot() +
  facet_wrap(~ trait, scales = "free", ncol = 4, nrow = 2)
```

```{r}
ITV_w %>% 
  ggplot(aes(x = landscape, y = ITV, color = habitat)) + 
  geom_point() +
  facet_wrap( ~ trait, scales = "free", nrow = 2, ncol = 4) + 
  theme_classic() + 
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.margin = margin(t = -8),
        strip.background = element_rect(color = "black", size = 1))
```

#### fixed = habitat
```{r}
hbt <- c("fen", "heath", "ridge")
fix_avg_hbt <- data.frame()

for(h in 1:length(hbt)) {
  spec_traits_bs <- read.csv("2_data/2_clean/traits/filled_traits_bmWeighted.csv") %>%
  filter(habitat == hbt[h]) %>% 
  group_by(spec_comm, trait) %>% 
  summarise(trait_value = mean(trait_value)) %>% 
  rename(spec = spec_comm) %>% 
  pivot_wider(names_from = trait, values_from = trait_value)
  
x <- community %>% 
  left_join(spec_traits_bs) %>%
  filter(part == "leaf", 
         habitat == hbt[h]) %>% 
  mutate(height = biomass * height, 
         LDMC = biomass * LDMC, 
         SLA = biomass * SLA, 
         wN = biomass * wN, 
         d15N = biomass * d15N, 
         wC = biomass * wC, 
         d13C = biomass * d13C, 
         CN_ratio = biomass * CN_ratio) %>% 
  group_by(plot) %>% 
  summarise(biomass = sum(biomass, na.rm = TRUE), 
            height = sum(height, na.rm = TRUE), 
            LDMC = sum(LDMC, na.rm = TRUE), 
            SLA = sum(SLA, na.rm = TRUE), 
            wN = sum(wN, na.rm = TRUE), 
            d15N = sum(d15N, na.rm = TRUE), 
            wC = sum(wC, na.rm = TRUE), 
            d13C = sum(d13C, na.rm = TRUE), 
            CN_ratio = sum(CN_ratio, na.rm = TRUE)) %>% 
  mutate(var_type = "landscape", 
         habitat = hbt[h],
         height = height / biomass, 
         LDMC = LDMC / biomass, 
         SLA = SLA / biomass, 
         wN = wN / biomass, 
         d15N = d15N / biomass, 
         wC = wC / biomass, 
         d13C = d13C / biomass, 
         CN_ratio = CN_ratio / biomass
         ) %>% 
  select(!biomass)

if (nrow(fix_avg_hbt) == 0) {
     fix_avg_hbt <- x
     } else {
       fix_avg_hbt <- bind_rows(fix_avg_hbt, x)
     }
}

```

```{r}
ITV_hbt <- bind_rows(fix_avg, spec_avg) %>% 
  select(!c(habitat, biomass)) %>% 
  pivot_longer(cols = height:CN_ratio, names_to = "trait", values_to = "trait_value") %>% 
  pivot_wider(names_from = var_type, 
              values_from = trait_value) %>% 
  mutate(ITV = local - landscape) %>% 
  pivot_longer(cols = landscape:ITV, names_to = "var_type", values_to = "CWM") %>% 
  mutate(habitat = as.factor(ifelse( # create habitat variable based on plot code
    grepl('D', plot), 'fen', ifelse(
      grepl('E', plot), 'heath', 'ridge'))
    )) %>% 
  mutate(treatment = as.factor(ifelse( # create treatment variable based on plot code
    grepl('C', plot), 'control', 'snowfence'
    ))) %>%  
  mutate(site_id = as.factor(substr(plot, 1,2))) %>%  # create site variable
  select(plot, site_id, habitat, treatment, everything())
```

```{r}
ITV_hbt_w <- ITV_hbt %>% 
  pivot_wider(names_from = var_type, values_from = CWM)

ITV_hbt %>% 
  filter(var_type == "local") %>% 
  ggplot(aes(x = habitat, y = CWM, fill = treatment)) + 
  geom_boxplot() +
  facet_wrap(~ trait, scales = "free", ncol = 4, nrow = 2)
```

```{r}
ITV_hbt_w %>% 
  ggplot(aes(x = landscape, y = ITV, color = habitat)) + 
  geom_point(aes(shape = treatment)) +
  facet_wrap( ~ trait, scales = "free", nrow = 2, ncol = 4) + 
  theme_classic() + 
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.margin = margin(t = -8),
        strip.background = element_rect(color = "black", size = 1))
```

### model test
```{r}
# traits are normally distributed it seems
spec_avg %>% 
  pull(wC) %>% 
  fitdist("norm") %>% 
  plot()
```

```{r}
# fixed average anova
fit <- ITV %>% 
  filter(trait == "CN_ratio", var_type == "landscape") %>% 
  lm(CWM ~ habitat * treatment, data = .)

summary(fit)
model_ouptut <- car::Anova(fit, type = "II") %>% 
  tidy() %>% 
  mutate(var_type = "landscape") %>% 
  select(var_type, everything())

# specific average anova
fit <- ITV %>% 
  filter(trait == "CN_ratio", var_type == "local") %>% 
  lm(CWM ~ habitat * treatment, data = .)

summary(fit)
model_ouptut <- car::Anova(fit, type = "II") %>% 
  tidy() %>% 
  mutate(var_type = "local") %>% 
  select(var_type, everything()) %>% 
  bind_rows(model_ouptut, .)

# specific average anova
fit <- ITV %>% 
  filter(trait == "CN_ratio", var_type == "ITV") %>% 
  lm(CWM ~ habitat * treatment, data = .)

model_ouptut <- car::Anova(fit, type = "II") %>% 
  tidy() %>% 
  mutate(var_type = "ITV") %>% 
  select(var_type, everything()) %>% 
  bind_rows(model_ouptut, .)

var_test <- model_ouptut %>% 
  select(var_type:sumsq) %>% 
  pivot_wider(id_cols = term, names_from = var_type, values_from = sumsq) %>% 
  rename(turnover = landscape, 
         total = local) %>% 
  mutate(covariation = total - (turnover + ITV)) %>% 
  select(term, turnover, ITV, covariation, total)

# get variability 
var_test <- var_test %>%
  # Summarise every column that is a number
  summarise(
    across(turnover:total, sum, na.rm = TRUE),
    # Manually add a label for the new row (essential for bind_rows)
    term = "total" 
  ) %>% 
  bind_rows(var, .)

# get variablility proportions
var_prop_test <- var_test %>% 
  filter(term == "total") %>% 
  pull(total) %>%
  {total_val <- .; mutate(var, across(where(is.numeric), ~ .x / total_val))} %>% 
  mutate(trait = "CN_ratio", .before = "term")
```

### model loop
#### fixed = landscape
```{r}
# loop prep
tr <- unique(ITV$trait)
vt <- unique(ITV$var_type) 

vars <- data.frame()
var_prop <- data.frame()
i <- 1
j <- 1

ITV_models <- list()

for(i in 1:length(tr)) {
  key <- tr[i]
  
  model_data <- ITV %>%
    filter(trait == tr[i])
  
  model_output <- data.frame() # create model output place holder
  
  for(j in 1:length(vt)) {
    model_data_lvl2 <- model_data %>% 
      filter(var_type == vt[j])
    
    model_key <- paste(tr[i], vt[j], sep = "_")
    
    fit <- lm(CWM ~ habitat * treatment, data = model_data_lvl2)
    ITV_models[[model_key]] <- fit
     
    x <- car::Anova(fit, type = "II") %>% 
      tidy() %>% 
      mutate(var_type = vt[j]) %>% 
      select(var_type, everything())
    
    if (nrow(model_output) == 0) {
      model_output <- x
    } else {
        model_output <- bind_rows(model_output, x)
      }
  }
  
  
  
  var <- model_output %>% 
    select(var_type:sumsq) %>% 
    pivot_wider(id_cols = term, names_from = var_type, values_from = sumsq) %>% 
    rename(turnover = landscape, 
           total = local) %>% 
    mutate(covariation = total - (turnover - ITV)) %>% 
    select(term, turnover, ITV, covariation, total)
  
  # get variability 
  var <- var %>%
    # Summarise every column that is a number
    summarise(
      across(turnover:total, sum, na.rm = TRUE),
      # Manually add a label for the new row (essential for bind_rows)
      term = "total" 
    ) %>% 
    bind_rows(var, .) %>% 
    mutate(trait = tr[i], .before = "term")
  
  # get sums of squares
  if (nrow(vars) == 0) {
      vars <- var
      } else {
        vars <- bind_rows(vars, var)
      }
  
  # get variablility proportions
  x <- var %>% 
    filter(term == "total") %>% 
    pull(total) %>%
    {total_val <- .; mutate(var, across(where(is.numeric), ~ .x / total_val))} 
  
  if (nrow(var_prop) == 0) {
      var_prop <- x
      } else {
        var_prop <- bind_rows(var_prop, x)
    }
}

# cleaning up
rm(model_key, key, x, i, j, tr, vt)
```

```{r}
# get model statistics (goodness of fit)
model_fit <- map_dfr(names(ITV_models), function(tr_vt) {
  model <- ITV_models[[tr_vt]]
  
  # Get R-squared
  mod_stats <- glance(model) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
  
  # Combine with species name
  bind_cols(var_trait = tr_vt, mod_stats)
})

# get hypothesis tests
hyp_test <- map_dfr(names(ITV_models), function(tr_vt) {
  model <- ITV_models[[tr_vt]]
  
# Get model anovas
  test <- car::Anova(model, type = "II") %>% 
    tidy() %>% 
    mutate(var_trait = tr_vt) %>% 
    dplyr::select(var_trait, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})

# pairwise comparisons 
pair_comps <- map_dfr(names(ITV_models), function(tr_vt) {
  model <- ITV_models[[tr_vt]]
  
  # Get R-squared
  emm <- emmeans(model, ~ treatment * habitat)
  pair_comp <- pairs(emm, adjust = "tukey") %>%
    as.data.frame() %>% 
    mutate(var_trait = tr_vt) %>% 
    dplyr::select(var_trait, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})
```

```{r}
# write results to excel file
wb <- createWorkbook()
addWorksheet(wb, "model_fit")
writeData(wb, "model_fit", model_fit)

addWorksheet(wb, "hypothesis_test")
writeData(wb, "hypothesis_test", hyp_test)

addWorksheet(wb, "pair_comparison")
writeData(wb, "pair_comparison", pair_comps)

addWorksheet(wb, "sumsq")
writeData(wb, "sumsq", vars)

addWorksheet(wb, "variance_proportions")
writeData(wb, "variance_proportions", var_prop)

# Save the workbook
saveWorkbook(wb, "3_output/2_statistics/traits/ITV_lm.xlsx", overwrite = TRUE)
```

```{r}
var_prop %>% 
  filter(term == "total") %>% 
  select(!total) %>% 
  pivot_longer(cols = turnover:covariation, names_to = "var_type", values_to = "variance") %>% 
  filter(variance >= 0) %>%
  ggplot(aes(x = trait, y = variance, fill = var_type)) +
  geom_col(position = "fill") + 
  scale_y_continuous(expand = c(0,0)) + 
  theme_classic() +
  theme(legend.position = "bottom",
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "cm"))
```

#### fixed = habitat
```{r}
# loop prep
hbt <- unique(ITV_hbt$habitat)
tr <- unique(ITV_hbt$trait)
vt <- unique(ITV_hbt$var_type) 

vars <- data.frame()
var_prop <- data.frame()
ITV_models <- list()

for(h in 1:length(hbt)) {
  ITV_dat <- ITV_hbt %>% 
    filter(habitat == hbt[h])
  
  for(i in 1:length(tr)) {
  model_data <- ITV_dat %>%
    filter(trait == tr[i])
  
  model_output <- data.frame() # create model output place holder
  
    for(j in 1:length(vt)) {
      model_data_lvl2 <- model_data %>% 
        filter(var_type == vt[j])
      
      model_key <- paste(hbt[h], tr[i], vt[j], sep = "_")
      
      fit <- lm(CWM ~ treatment, data =   model_data_lvl2)
      ITV_models[[model_key]] <- fit
       
      x <- car::Anova(fit, type = "II") %>% 
        tidy() %>% 
        mutate(var_type = vt[j]) %>% 
        select(var_type, everything())
      
      if (nrow(model_output) == 0) {
        model_output <- x
      } else {
          model_output <- bind_rows(model_output, x)
        }
    }
    
 
 var <- model_output %>% 
   select(var_type:sumsq) %>% 
   pivot_wider(id_cols = term, names_from = var_type,   values_from = sumsq) %>% 
   rename(turnover = landscape, 
          total = local) %>% 
   mutate(covariation = total - (turnover + ITV)) %>% 
   select(term, turnover, ITV, covariation, total)
 
 # get variability 
 var <- var %>%
   # Summarise every column that is a number
   summarise(
     across(turnover:total, sum, na.rm = TRUE),
     # Manually add a label for the new row (essential f bind_rows)
     term = "total" 
   ) %>% 
   bind_rows(var, .) %>% 
   mutate(habitat = hbt[h], 
          trait = tr[i], 
          .before = "term")
 
 # get sums of squares
 if (nrow(vars) == 0) {
     vars <- var
     } else {
       vars <- bind_rows(vars, var)
     }
 
 # get variablility proportions
 x <- var %>% 
   filter(term == "total") %>% 
   pull(total) %>%
   {total_val <- .; mutate(var, across(where(is.numeric), ~ .x / total_val))} 
 
 if (nrow(var_prop) == 0) {
     var_prop <- x
     } else {
       var_prop <- bind_rows(var_prop, x)
   }
  }
}

# cleaning up
rm(model_key, x, i, j, tr, vt, hbt)
```

```{r}
# get model statistics (goodness of fit)
model_fit <- map_dfr(names(ITV_models), function(tr_vt) {
  model <- ITV_models[[tr_vt]]
  
  # Get R-squared
  mod_stats <- glance(model) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
  
  # Combine with species name
  bind_cols(var_trait = tr_vt, mod_stats)
})

# get hypothesis tests
hyp_test <- map_dfr(names(ITV_models), function(tr_vt) {
  model <- ITV_models[[tr_vt]]
  
# Get model anovas
  test <- car::Anova(model, type = "II") %>% 
    tidy() %>% 
    mutate(var_trait = tr_vt) %>% 
    dplyr::select(var_trait, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})

# pairwise comparisons 
pair_comps <- map_dfr(names(ITV_models), function(tr_vt) {
  model <- ITV_models[[tr_vt]]
  
  # Get R-squared
  emm <- emmeans(model, ~ treatment)
  pair_comp <- pairs(emm, adjust = "tukey") %>%
    as.data.frame() %>% 
    mutate(var_trait = tr_vt) %>% 
    dplyr::select(var_trait, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})
```

```{r}
# write results to excel file
wb <- createWorkbook()
addWorksheet(wb, "model_fit")
writeData(wb, "model_fit", model_fit)

addWorksheet(wb, "hypothesis_test")
writeData(wb, "hypothesis_test", hyp_test)

addWorksheet(wb, "pair_comparison")
writeData(wb, "pair_comparison", pair_comps)

addWorksheet(wb, "sumsq")
writeData(wb, "sumsq", vars)

addWorksheet(wb, "variance_proportions")
writeData(wb, "variance_proportions", var_prop)

# Save the workbook
saveWorkbook(wb, "3_output/2_statistics/traits/ITV_hbt_lm.xlsx", overwrite = TRUE)
```

```{r}
vars %>% 
  filter(term == "total") %>% 
  mutate(turnover_itv = turnover + ITV)
```

```{r}
var_prop %>% 
  filter(term == "total") %>% 
  select(!total) %>% 
  pivot_longer(cols = turnover:covariation, names_to = "var_type", values_to = "variance") %>% 
  ggplot(aes(x = trait, y = variance, fill = var_type)) +
  geom_col(position = "fill") + 
  facet_wrap(~ habitat, nrow = 3)
```

### trait.flex
```{r}
x <- spec_avg %>% 
  select(plot, wC) %>% 
  rename(spec.avg = wC) 

x <- fix_avg %>% 
  select(wC) %>% 
  rename(fix.avg = wC) %>% 
  bind_cols(x, .) %>% 
  mutate(habitat = as.factor(ifelse( # create habitat variable based on plot code
    grepl('D', plot), 'fen', ifelse(
      grepl('E', plot), 'heath', 'ridge'))
    )) %>% 
  mutate(treatment = as.factor(ifelse( # create treatment variable based on plot code
    grepl('C', plot), 'control', 'snowfence'
    ))) %>%  
  mutate(site_id = as.factor(substr(plot, 1,2))) 

x1 <- trait.flex.anova(~ habitat * treatment, spec.avg, fix.avg, data = x)
x1[["RelSumSq"]]
anova(ITV_models[["fen_CN_ratio_landscape"]])

plot.trait.flex(x1, plot.covar=TRUE, use.percentage=T)
```
### varpart* 
#### fixed = landscape
```{r}
tr <- spec_avg %>% # CWM local
  select(!var_type) %>% 
  column_to_rownames("plot") %>%
  colnames()

v <- c("turnover", "ITV", "covariation")
varpart_r2 <- data.frame()

for(t in 1:length(tr)) {
  y <- spec_avg %>% # CWM local
    select(!var_type) %>% 
    column_to_rownames("plot") %>% 
    select(t)
  
  x1 <- fix_avg %>%  # CWM landscape
    select(!c(var_type, biomass)) %>% 
    column_to_rownames("plot") %>% 
    relocate(tr) %>% 
    select(t)
  
  x2 <- x1 - y # ITV
    
  mod <- varpart(y, x1, x2)
  x <- mod[["part"]][["indfract"]] %>% 
    slice_head(n = 3) %>% 
    pull(Adj.R.squared) %>% 
    data.frame(trait = tr[t], r2 = ., var_part = as.factor(v))
  
  if (nrow(varpart_r2) == 0) {
      varpart_r2 <- x
      } else {
        varpart_r2 <- bind_rows(varpart_r2, x)
      }
  
  rm(y, x1, x2, x) # cleaning up
}

write.csv(varpart_r2, "3_output/2_statistics/traits/varpart_r2.csv")
```

```{r}
varpart_r2 <- varpart_r2 %>% mutate(trait = factor(trait, levels = c("height", "SLA", "LDMC", "CN_ratio", "wC", "wN", "d13C", "d15N")), var_part = factor(var_part, levels = c("ITV", "covariation", "turnover")))

trait_labs <- c("Plant~height", "SLA", "LDMC", "C:N~ratio", "'%'~C", "'%'~N", "delta^13*C", "delta^15*N")
names(trait_labs) <- c("height", "SLA", "LDMC", "wC", "wN", "CN_ratio", "d13C", "d15N")

varpart_r2 %>% 
  filter(r2 >= 0) %>%
  ggplot(aes(x = trait, y = r2, fill = var_part)) +
  geom_col(position = "fill") + 
  scale_x_discrete(labels = parse(text = trait_labs)) + 
  scale_y_continuous(expand = c(0,0)) + 
  theme_classic() +
  theme(legend.position = "bottom",
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "cm")) + 
  coord_flip()


```

#### fixed = habitat
```{r}
tr <- spec_avg %>% # CWM local
  select(!var_type) %>% 
  column_to_rownames("plot") %>%
  colnames()

v <- c("turnover", "ITV", "covariation")
varpart_r2_hbt <- data.frame()

hbt <- c("fen", "heath", "ridge")

for(h in 1:length(hbt)) {
  y_hbt <- spec_avg %>% 
    mutate(habitat = as.factor(ifelse( # create habitat variable based on plot code
    grepl('D', plot), 'fen', ifelse(
      grepl('E', plot), 'heath', 'ridge'))
    )) %>% 
    filter(habitat == hbt[h]) %>% # CWM local
    select(!var_type, habitat) %>% 
    column_to_rownames("plot") 
  
  x1_hbt <- fix_avg_hbt %>%  # CWM landscape
    filter(habitat == hbt[h]) %>% 
    select(!c(var_type, biomass, habitat)) %>% 
    column_to_rownames("plot") %>% 
    relocate(tr) 
    
   
  for(t in 1:length(tr)) {
    y <- y_hbt %>% 
      select(t)
    
    x1 <- x1_hbt %>% 
      select(t)
    
    x2 <- x1 - y # ITV
      
    mod <- varpart(y, x1, x2)
    x <- mod[["part"]][["indfract"]] %>% 
      slice_head(n = 3) %>% 
      pull(Adj.R.squared) %>% 
      data.frame(habitat = hbt[h], 
                 trait = tr[t], 
                 r2 = ., 
                 var_part = as.factor(v))
    
    if (nrow(varpart_r2_hbt) == 0) {
        varpart_r2_hbt <- x
        } else {
          varpart_r2_hbt <- bind_rows(varpart_r2_hbt, x)
        }
  }
  
  rm(y, y_hbt, x1, x1_hbt, x2, x) # cleaning up
}

write.csv(varpart_r2_hbt, "3_output/2_statistics/traits/varpart_r2_hbt.csv")
```

```{r}
varpart_r2_hbt <- varpart_r2_hbt %>% mutate(trait = factor(trait, levels = c("height", "SLA", "LDMC", "CN_ratio", "wC", "wN", "d13C", "d15N")), var_part = factor(var_part, levels = c("ITV", "covariation", "turnover")))

trait_labs <- c("Plant~height", "SLA", "LDMC", "C:N~ratio", "'%'~C", "'%'~N", "delta^13*C", "delta^15*N")
names(trait_labs) <- c("height", "SLA", "LDMC", "wC", "wN", "CN_ratio", "d13C", "d15N")

varpart_r2_hbt %>% 
  filter(r2 >= 0) %>%
  ggplot(aes(x = trait, y = r2, fill = var_part)) +
  geom_col(position = "fill") + 
  scale_x_discrete(labels = parse(text = trait_labs)) + 
  scale_y_continuous(expand = c(0,0)) + 
  theme_classic() +
  theme(legend.position = "bottom",
        legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "cm")) + 
  facet_wrap(~habitat)


```

## local vs. global
```{r}
try_traits <- try_traits %>% 
  mutate(treatment = "global") %>% 
  left_join(spec_list) %>% 
  mutate(across(c(1:3, 5, 6), as.factor))

all_traits <- bind_rows(try_traits, traits_l)
```

```{r}
facet_order <- c("height", "SLA", "LDMC", "CN_ratio", "wC", "wN", "d13C", "d15N")
trait_labs <- c("Plant~height", "SLA", "LDMC", "C:N~ratio", "'%'~C", "'%'~N", "delta^13*C", "delta^15*N")
names(trait_labs) <- facet_order


all_traits %>% 
  filter(#spec == "Betnan",
         trait %in% c("SLA", "LDMC", "height", "wC", "wN", "CN_ratio", "d13C", "d15N")) %>% 
  mutate(trait = factor(trait, levels = facet_order), 
         treatment = factor(treatment, levels = c("global", "control", "snowfence"))) %>%
  ggplot(aes(x = group, y = trait_value, fill = treatment)) +
  #geom_violin(trim = FALSE, position = position_dodge(.8),) +
  geom_boxplot(outliers = FALSE, position = position_dodge(.8), width = .6) +
  geom_jitter(aes(color = treatment), alpha = 0.2,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8)) +
  scale_fill_manual(values = c("#ff6666", "#999999", "#4da6ff")) +
  scale_color_manual(values = c("#ff6666", "#999999", "#4da6ff")) +
  facet_wrap(~ trait, scales = "free", ncol = 4, nrow = 2,
             labeller = as_labeller(trait_labs, label_parsed)) +
  scale_x_discrete(labels=c("dS" = "d. shrubs", "eS" = "e. shrubs", "F" = "forbs", "G" = "graminoids")) +
  labs(x = "Functional group", y = "Trait value", fill = "Dataset", color = "Dataset") +
  theme_classic() +
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.margin = margin(t = -8),
        strip.background = element_rect(color = "black", size = 1),
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/1_figures/traits/local_global_traits_comparison.tif", height = 5, width = 8)
```

### model test
```{r seedling count distribution, echo = FALSE}
d <- all_traits %>% 
  filter(spec == "Empnig" & trait == "wN") 
dist <- fitdist(d$trait_value, "norm")
plot(dist)
```

```{r}
# lmm (produces singular warning!)
fit1 <- lm(trait_value ~ treatment, data = filter(all_traits, spec == "Empnig" & trait == "height"))

summary(fit1)
car::Anova(fit1, type = "II")

emmeans(fit1, pairwise ~ treatment) %>% 
  pairs() %>% 
  as.data.frame() 
```

```{r}
# model with covariate adjustment
fit2 <- lm(trait_value ~ treatment * habitat, data = filter(traits_l, spec == "Empnig" & trait == "height"))

summary(fit2)
car::Anova(fit2, type = "II") 

glance(fit2)

AIC(fit1, fit2)
```

```{r trait model diagnostics}
sim_resid <- simulateResiduals(fittedModel = fit1)
plot(sim_resid)
testDispersion(sim_resid); rm(sim_resid)

check_singularity(fit1, tolerance = 1e-05)


x <- check_collinearity(fit1)
plot(x); rm(x)
```

### model loop
```{r}
# exclude species with insufficient data
model_sp_traits <- all_traits %>% 
  group_by(group, trait) %>% 
  filter(n_distinct(treatment) >= 2) %>%
  filter(!trait %in% c("dweight", "fweight", "leafa"))

gr <- unique(model_sp_traits$group)
tr <- unique(model_sp_traits$trait) 
  
group_trait_grid <- expand_grid(group = gr, trait = tr) %>% 
  mutate(group = as.character(group)) %>% 
  as.data.frame()
model_list <- list()
i <- 1
```

```{r}
for(i in 1:nrow(group_trait_grid)) {
  key <- paste(group_trait_grid[i, "group"], group_trait_grid[i, "trait"], sep = "_")
  model_data <- all_traits %>% 
    filter(group == group_trait_grid[i, "group"] & 
             trait == group_trait_grid[i, "trait"])
  model_list[[key]] <- lm(trait_value ~ treatment, data = model_data)
  rm(model_data, key)
}

# View a specific model
summary(model_list[["dS_wN"]])
```


```{r}
# extract model coefficients
get_model_pars <- function(grtr, model_list) {
  model <- model_list[[grtr]]
  
  # Get coefficients (estimate, std.error, p.value)
  p.vals <- car::Anova(model, type = "II") %>% 
  tidy() %>% 
  dplyr::select(term, p.value) %>% 
  pivot_wider(names_from = term, values_from = p.value) %>% 
  dplyr::select(!Residuals)
  
  # Get R-squared
  model_pars <- glance(model) %>% 
    dplyr::select(r.squared:p.value) 
  
  # Combine with species name
  bind_cols(spec = grtr, model_pars, p.vals)
}

results <- map(names(model_list), safely(get_model_pars), model_list = model_list)
results_list <- map(results, "result")
results_df <- results_list[!map_lgl(results_list, is.null)] %>%
  list_rbind()

results_df <- results_df %>% 
  mutate(sig = ifelse(p.value < 0.05, "*", "ns"))
```

```{r}
# get model statistics (goodness of fit)
model_fit <- map_dfr(names(model_list), function(gr_tr) {
  model <- model_list[[gr_tr]]
  
  # Get R-squared
  mod_stats <- glance(model) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
  
  # Combine with species name
  bind_cols(group_trait = gr_tr, mod_stats)
})

# get hypothesis tests
hyp_test <- map_dfr(names(model_list), function(gr_tr) {
  model <- model_list[[gr_tr]]
  
# Get model anovas
  test <- car::Anova(model, type = "II") %>% 
    tidy() %>% 
    mutate(group_trait = gr_tr) %>% 
    dplyr::select(group_trait, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})

# pairwise comparisons 
pair_comps <- map_dfr(names(model_list), function(gr_tr) {
  model <- model_list[[gr_tr]]
  
  # Get R-squared
  emm <- emmeans(model, ~ treatment)
  pair_comp <- pairs(emm, adjust = "tukey") %>%
    as.data.frame() %>% 
    mutate(group_trait = gr_tr) %>% 
    dplyr::select(group_trait, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})
```

```{r}
# write results to excel file
wb <- createWorkbook()
addWorksheet(wb, "model_fit")
writeData(wb, "model_fit", model_fit)

addWorksheet(wb, "hypothesis_test")
writeData(wb, "hypothesis_test", hyp_test)

addWorksheet(wb, "pair_comparison")
writeData(wb, "pair_comparison", pair_comps)

# Save the workbook
saveWorkbook(wb, "3_output/2_statistics/traits/local_global_trait_comp_lm.xls", overwrite = TRUE)
```


## species response to treatment
How does individual species response differ between treatments? what is the direction of this response and what is it related to? 
```{r plant height plot, echo = FALSE}
traits %>% 
  ggplot(aes(x = habitat, y = height, color = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~ spec) +
  theme(legend.position = c(.85, .8), 
        axis.text.x = element_text(angle = 25, hjust = 1))

ggsave("3_output/Figures/plant_height.svg", height = 4, width = 25)
```

### model test
```{r seedling count distribution, echo = FALSE}
d <- traits_l %>% 
  filter(spec == "Empnig" & trait == "height") 
dist <- fitdist(d$trait_value, "norm")
plot(dist)
```

```{r}
# lmm (produces singular warning!)
fit1 <- lmer(trait_value ~ treatment + (1 |habitat), data = filter(traits_l, spec == "Empnig" & trait == "height"))

summary(fit1)
car::Anova(fit1, type = "II")

x <- emmeans(fit1, pairwise ~ treatment) %>% 
  pairs() %>% 
  as.data.frame() 
```

```{r}
# model with covariate adjustment
fit2 <- lm(trait_value ~ treatment * habitat, data = filter(traits_l, spec == "Empnig" & trait == "height"))

summary(fit2)
car::Anova(fit2, type = "II") 

glance(fit2)

AIC(fit1, fit2)
```

```{r trait model diagnostics}
sim_resid <- simulateResiduals(fittedModel = fit2)
plot(sim_resid)
testDispersion(sim_resid); rm(sim_resid)

check_singularity(fit2, tolerance = 1e-05)


x <- check_collinearity(fit2)
plot(x); rm(x)
```

### model loop
```{r}
# exclude species with insufficient data
model_sp_traits <- traits_l %>% 
  group_by(spec, trait) %>% 
  filter(n_distinct(treatment) >= 2 & n_distinct(habitat) >= 2) %>%
  filter(!trait %in% c("dweight", "fweight", "leafa"))

sp <- unique(model_sp_traits$spec)
tr <- unique(model_sp_traits$trait) 
  
spec_trait_grid <- expand_grid(spec = sp, trait = tr) %>% 
  mutate(spec = as.character(spec)) %>% 
  as.data.frame()
model_list <- list()
i <- 1
```

```{r}
for(i in 1:nrow(spec_trait_grid)) {
  key <- paste(spec_trait_grid[i, "spec"], spec_trait_grid[i, "trait"], sep = "_")
  model_data <- traits_l %>% 
    filter(spec == spec_trait_grid[i, "spec"] & 
             trait == spec_trait_grid[i, "trait"])
  model_list[[key]] <- lm(trait_value ~ treatment * habitat, data = model_data)
  rm(model_data, key)
}

# View a specific model
summary(model_list[["Empnig_height"]])
```


```{r}
# extract model coefficients
get_model_pars <- function(sptr, model_list) {
  model <- model_list[[sptr]]
  
  # Get coefficients (estimate, std.error, p.value)
  p.vals <- car::Anova(model, type = "II") %>% 
  tidy() %>% 
  dplyr::select(term, p.value) %>% 
  pivot_wider(names_from = term, values_from = p.value) %>% 
  dplyr::select(!Residuals)
  
  # Get R-squared
  model_pars <- glance(model) %>% 
    dplyr::select(r.squared:p.value) 
  
  # Combine with species name
  bind_cols(spec = sptr, model_pars, p.vals)
}

results <- map(names(model_list), safely(get_model_pars), model_list = model_list)
results_list <- map(results, "result")
results_df <- results_list[!map_lgl(results_list, is.null)] %>%
  list_rbind()

results_df <- results_df %>% 
  mutate(sig = ifelse(p.value < 0.05, "*", "ns"))
```
```{r}
# get model statistics (goodness of fit)
model_fit <- map_dfr(names(model_list), function(gr) {
  model <- model_list[[gr]]
  
  # Get R-squared
  mod_stats <- glance(model) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
  
  # Combine with species name
  bind_cols(spec = gr, mod_stats)
})

# get hypothesis tests
hyp_test <- map_dfr(names(model_list), function(gr) {
  model <- model_list[[gr]]
  
  # Check residual df
  resid_df <- df.residual(model)
  
  # Skip if no residual df
  if (is.null(resid_df) || resid_df == 0) {
    warning(paste("Skipping", gr, "- residual df = 0"))
    return(NULL)
  }
  
  # Get model anovas
  test <- car::Anova(model, type = "II") %>% 
    tidy() %>% 
    mutate(group = gr) %>% 
    dplyr::select(group, everything()) %>% 
    mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    p.value < 0.1 ~ ".",
    TRUE ~ ""
  ))
})

# pairwise comparisons 
# Create safe version
safe_pairs <- possibly(
  function(model, gr) {
    emm <- emmeans(model, ~ treatment | habitat)
    pairs(emm, adjust = "tukey") %>%
      as.data.frame() %>% 
      mutate(group = gr) %>% 
      dplyr::select(group, everything()) %>% 
      mutate(significance = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01 ~ "**",
        p.value < 0.05 ~ "*",
        p.value < 0.1 ~ ".",
        TRUE ~ ""
      ))
  },
  otherwise = NULL
)

# Apply to all models
pair_comps <- map_dfr(names(model_list), function(gr) {
  model <- model_list[[gr]]
  safe_pairs(model, gr)
}, .progress = TRUE)
```

```{r}
# write results to excel file
wb <- createWorkbook()
addWorksheet(wb, "model_fit")
writeData(wb, "model_fit", model_fit)

addWorksheet(wb, "hypothesis_test")
writeData(wb, "hypothesis_test", hyp_test)

addWorksheet(wb, "pair_comparison")
writeData(wb, "pair_comparison", pair_comps)

# Save the workbook
saveWorkbook(wb, "3_output/2_statistics/traits/ITV_lm_results.xls", overwrite = TRUE)
```

### bayesian test
```{r}
# setting flat prior

brm(formula = height ~ treatment * habitat + (1 | site_id), data = traits, family = gaussian())
```

